---
title: "Practical Machine Learning Course Project"
author: "Jonathan Piper"
date: "23 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Qualitative Activity Recognition of Weight Lifting Exercises 

## Executive Summary

Researchers set out to measure how well subjects performed dumbell exercises and to provide feedback. 
Subjects performed exercises in each of 5 ways.


  A according to specification        
  B throwing elbows to the front      
  C lifting the dumbells only halfway 
  D lowering the dumbell only halfway 
  E throwing the hips to the front    

Data was derived from accelerometers on the belt, forearm, arm, and dumbell of the subjects. 

The goal of the assignment is to use the training data set of 19,622 obervations, to build a prediction
model to classify accurately the 20 test data between categorries A to E.


### Read in the traing and test data

```{r read data}
url_train<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_test<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training_zero<- read.csv(url_train)
test<- read.csv(url_test)
```


### Take a look at the training data

``` {r}
head(training_zero)
head(training_zero)
str(training_zero)
apply(training_zero,2,function(x){sum(is.na(x))})
``` 

### Clean Data

NA's and "" values - it is clear that data columns either have no NA's or "" values, or many (always 19216 out of xxxxxx).
Columns 1 through 7 have no predictive value - timestamp, new window, sbject name etc

```{r delete NAs,"",first 7 cols}
library(dplyr)
training1<-training_zero[,!apply(training_zero,2,function(x) any(is.na(x)))]
training2<-training1[,!apply(training1,2,function(x){sum((x==""))})>=1]
training3<- select(training2,-c(1,2,3,4,5,6,7))
```

All predictor variable data is now numeric or intger, free of mssing values and NAs.

### Notes on method and approach

Reviewing discussion forum, particularly Len Greski's tutorials indicates that:-
* a high degree of predictive accuracy is required in order to achive reliable predicions on every element of the twenty row test set 
* cross validation is necesssary in order to assess theh accuracy   
* parallel processing is essential ot get sensible run times

I selected a random forest, as this algorithm is known to be accurate.  The downside of this method is lack of interpreatbility, but 
for this assignment interpretabiliy is not a priority.

For a random forest method, any further data pre-processing, such as scaling, centring, or otherwise transforming the data is unnecessary. 

The choice of model was reinforced following a reading of the original paper Qualitative Activity Recognition of Weight Lifting Exercises.  

### Build and Run the Model

Configure parallel processing

```{r}
library(parallel)
library(doParallel)

cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

Training Model

```{r}
library(caret)
set.seed <- 4321 
start.time <- Sys.time()
fit <- train(classe ~ ., data=training3, method="rf", trControl=trainControl(method="cv",number =5,allowParallel=TRUE), tuneGrid=data.frame(mtry=6))
end.time=Sys.time()
elapsed.time<- end.time - start.time
```

De-register parallel processing cluster

```{r}
stopCluster(cluster)
registerDoSEQ()
```


Final Model

```{r}
fit$finalModel
```

OOB (Out of Bag) error rate 0.28%, so accuracy 99.72%, sufficient to apply to test set. 

Apply to test set

```{r}
pred <- predict(fit, test) 
pred
```

Predicted values were submitted to the Quiz and all proved correct.

